---
title: "Comparing Dual and Triple HIV-1 Reverse Transcriptase Inhibitor Regimens on CD4 Counts in AIDS Patients: A Longitudinal Analysis"
format: html
editor: visual
---

```{r echo=FALSE}
# Load necessary libraries
library(labelled)   # labeling data
library(rstatix)    # summary statistics
library(ggpubr)     # convenient summary statistics and plots
library(GGally)     # advanced plots
library(car)        # useful for ANOVA/Wald tests
library(Epi)        # easy CI for model coefficients/predictions
library(lme4)       # linear mixed-effects models
library(lmerTest)   # tests for linear mixed-effects models
library(emmeans)    # marginal means
library(multcomp)   # CI for linear combinations of model coefficients
library(geepack)    # generalized estimating equations
library(ggeffects)  # marginal effects, adjusted predictions
library(gt)         # nice tables
library(tidyverse)  # for data manipulation, visualization, and more
library(readr)      # to read CSV files
library(nlme)       # for mixed-effects models
library(reshape2)   # data reshaping
library(ggplot2)    # for data visualization
library(tidyr)      # for tidying data
library(dplyr)      # data manipulation
library(kableExtra) # enhanced tables
library(corrplot)   # for correlation matrix visualization
library(joineR)     # joint modeling of longitudinal and survival data
library(MASS)       # for statistical functions
library(lattice)    # for trellis graphics
library(gtsummary)  # for summary tables

# Set the theme for ggplot
theme_set(theme_minimal() + theme(legend.position = "bottom"))

# Load the data
actg <- read_csv("actg193a (1).csv")
head(actg)
```

# Introduction

The data are from a randomized, double-blind, study of AIDS patients with

advanced immune suppression (CD4 counts of less than or equal to 50 cells/mm\^3).  Patients in AIDS Clinical Trial Group (ACTG) Study 193A were randomized to dual or triple combinations of HIV-1 reverse transcriptase inhibitors.  Specifically, patients were randomized to one of four daily regimens containing 600mg of zidovudine: zidovudine alternating monthly with 400mg didanosine; zidovudine plus 2.25mg of zalcitabine; zidovudine plus 400mg of didanosine; or zidovudine plus 400mg of didanosine plus 400mg of nevirapine (triple therapy).

Measurements of CD4 counts were scheduled to be collected at baseline and at

8-week intervals during follow-up. However, the CD4 count data are unbalanced

due to mistimed measurements and missing data that resulted from skipped visits and dropout. The number of measurements of CD4 counts during the first 40 weeks of follow-up varied from 1 to 9, with a median of 4.

The response variable is the log transformed CD4 counts, log(CD4 counts + 1),

available on 1309 patients.

The categorical variable Treatment is coded

 The variable Week represents time since baseline (in weeks).

Variable List:

id: Subject ID

trt: Treatment (1 = zidovudine alternating monthly with 400mg didanosine, 2 = zidovudine plus 2.25mg of zalcitabine, 3 = zidovudine plus 400mg of didanosine, and 4 = zidovudine plus 400mg of didanosine plus 400mg of nevirapine.)

age: Age (years)

sex: Sex at birth (1=M, 0=F)

week: Week since baseline

logcd4: log(CD4 count + 1).

### Project Goal

# Methods

## Data Description

## Data Preparation

```{r echo=FALSE}
actg_new<-actg|>mutate_at(vars(trt, sex), as.factor)
actg_new$sex<-factor(actg_new$sex, levels = c("1", "0"), labels = c("M", "F"))
actg_new$trt <- fct_recode(actg_new$trt,
  "Zidovudine + Didanosine (Alternating)" = "1",
  "Zidovudine + Zalcitabine" = "2",
  "Zidovudine + Didanosine" = "3",
  "Zidovudine + Didanosine + Nevirapine" = "4"
)

```

## Exploratory Data Analysis

```{r echo=FALSE}
summary<-actg_new|>group_by(trt)|>get_summary_stats(logcd4, show=c("n", "min", "max", "mean", "sd"))
summary
```

## Statistical Analysis

# Results

# Discussion
