---
title: "Longitudinal Analysis"
author: "Obehi Winnifred Ikpea"
format: html
editor: visual
---

```{r echo=FALSE}
library(labelled)   # labeling data
library(rstatix)    # summary statistics
library(ggpubr)     # convenient summary statistics and plots
library(GGally)     # advanced plot
library(car)        # useful for anova/wald test
library(Epi)        # easy getting CI for model coef/pred
library(lme4)       # linear mixed-effects models
library(lmerTest)   # test for linear mixed-effects models
library(emmeans)    # marginal means
library(multcomp)   # CI for linear combinations of model coef
library(geepack)    # generalized estimating equations
library(ggeffects)  # marginal effects, adjusted predictions
library(gt)         # nice tables
library(tidyverse)  # for everything (data manipulation, visualization, coding, and more)
theme_set(theme_minimal() + theme(legend.position = "bottom")) # theme for ggplot
library(readr)
mcrf <- read_csv("mcrf (1).csv")
```

# Introduction


# Methods

## Data Preparation
Description:

The data are from the Muscatine Coronary Risk Factor (MCRF) study, 
a longitudinal survey of school-age children in Muscatine, Iowa.
The MCRF study had the goal of examining the development and
persistence of risk factors for coronary disease in children.
In the MCRF study, weight and height measurements of five cohorts of children, initially aged 5-7, 7-9, 9-11, 11-13, and 13-15 years, were
obtained biennially from 1977 to 1981. Data were collected on 4856
boys and girls. On the basis of a comparison of their weight to
age-gender specific norms, children were classified as obese or
not obese. 


Variable List:


id: participant ID
sex: sex at birth (0=Male,1=Female)
b_age: Baseline Age
age: Current Age
visit: visit number
obesity: Obesity Status (1=Obese, 0=Non-Obese, .=Missing)

```{r echo=FALSE}
head(mcrf)
mcrf_new<-mcrf|>mutate_at(vars(sex, obesity), as.factor)|>mutate_at(vars(visit), as.integer)
mcrf_new$sex<-factor(mcrf_new$sex, levels = c("0", "1"), labels=c("Male", "Female"))

head(mcrf_new)
```




## Exploratory Data Analysis
